---
title: "Reproducible Research: Peer Assessment 1"
author: "Pablo Rodríguez"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

We can unzip and load directly using unz:  

```{r}
data <- read.csv(unz("activity.zip", "activity.csv"))
```

It's always useful to take a look at the data:  

```{r}
head(data)
```


## What is mean total number of steps taken per day?

First of all, we exclude the missing values from our data:  

```{r}
cleanData <- data[!is.na(data$steps), ]
```

Create a new dataframe summarizing the total number of steps for each day:  

```{r}
stepsPerDay <- aggregate(cleanData$steps, by = list(cleanData$date), FUN = sum)
colnames(stepsPerDay) <- c("date", "steps")

head(stepsPerDay)
```

With this clean data, we can easily create an histogram:  

```{r}
barplot(stepsPerDay$steps, xlab = 'Day', ylab = 'Steps', main = 'Steps per day', names.arg = stepsPerDay$date)
```

Finally, calculate the mean:  

```{r}
mean <- mean(stepsPerDay$steps)
mean
```

And the median:  

```{r}
median <- median(stepsPerDay$steps)
median
```

## What is the average daily activity pattern?

Extract the daily data and plot it:  

```{r}
dailySteps <- aggregate(cleanData$steps, by = list(cleanData$interval), FUN = mean)
colnames(dailySteps) <- c("interval", "mean")
plot(dailySteps, type = "l", xlab = "interval", ylab = "frequency")
```

Extract the maximum value:  

```{r}
maxIndex <- which.max(dailySteps$mean)
maxInterval <- dailySteps[maxIndex, 1]
maxInterval
```


## Imputing missing values

We need to count the missing values:  

```{r}
numMissing <- sum(is.na(data$steps))
numMissing
```

Replace the nans with the means per interval across all days:  

```{r}
filledIn <- data
filledIn <- merge(filledIn, dailySteps) # This appends the means per interval

nans <- is.na(filledIn$steps) # Substitute the nans with means per interval
filledIn[nans, 2] <- filledIn[nans, 4] 
```

Take a look at the filled data:  

```{r}
stepsPerDayFilled <- aggregate(filledIn$steps, by = list(filledIn$date), FUN = sum)
colnames(stepsPerDayFilled) <- c("date", "steps")
barplot(stepsPerDayFilled$steps, xlab = 'Day', ylab = 'Steps', main = 'Steps per day (filled missing values)', names.arg = stepsPerDayFilled$date)
```

## Are there differences in activity patterns between weekdays and weekends?

Extract the day of the week: 

```{r}
dayName <- weekdays(as.Date(filledIn$date))
```

And append a new column containing information about the weekend status of each day:  

```{r}
weekend <- dayName == "sábado" | dayName == "domingo" # Please note I work on a spanish pc
filledIn <- cbind(filledIn, weekend)
```

Finally, plot the data for both cases:  

```{r}
weekdayData <- filledIn[filledIn$weekend == FALSE, ]
weekendData <- filledIn[filledIn$weekend == TRUE, ]

weekdayMean <- aggregate(weekdayData$steps, by = list(weekdayData$interval), FUN = mean)
weekendMean <- aggregate(weekendData$steps, by = list(weekendData$interval), FUN = mean)

colnames(weekdayMean) <- c("interval", "steps")
colnames(weekendMean) <- c("interval", "steps")

par(mfrow = c(2,1))
plot(weekdayMean$interval, weekdayMean$steps, type = 'l', xlab = "interval", ylab = "frequency", main = "Weekday")
plot(weekendMean$interval, weekendMean$steps, type = 'l', xlab = "interval", ylab = "frequency", main = "Weekend")
```

